%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Conclusion section
\section{Summary and Concluding Remarks}
\label{sec:conclusion}
This paper has studied a selection-on-observables approach to CM in a natural experiment setting.
I have shown the pitfalls of using the most popular methods for estimating direct and indirect effects without a clear case for the mediator being quasi-randomly assigned.
Using the Roy model as a benchmark, a mediator is unlikely to be quasi-randomly assigned in natural experiment settings, and the bias terms likely crowd out inference regarding CM effects.

This paper has also contributed to the growing CM literature in economics, connecting to MTE methods and developing a compelling way of estimating direct and indirect effects in a natural experiment setting.
It has also recognised limitations in the common practice of suggestive evidence for mechanisms, and given credible CM estimates in a famous natural experiment setting well-known in the economics field.
Further research could build on the approaches presented by suggesting efficiency improvements, adjustments for common statistical irregularities (say, cluster dependence), or integrating the MTE approach to the growing double robustness literature on CM \citep{farbmacher2022causal,bia2024double}.

These findings do not provide a blanket endorsement for applied researchers to use CM methods.
There are strong structural assumptions for adjusting identifying CM effects despite unobserved selection-into-mediator, and inference requires an IV for mediator take-up.
If these assumptions do not hold true, then selection-adjusted estimates of CM effects will also be biased, and will not improve on an unadjusted conventional approach.

Yet, there are likely settings in which the structural assumptions are credible.
Mediator monotonicity aligns well with economic theory in many cases, and it is plausible for researchers to study big data settings with external variation in mediator take-up costs.
In these cases, this paper opens the door to identifying mechanisms behind treatment effects in natural experiment settings.
% The approach could be used in AB tests, where a firm randomises a treatment and costs of a suspected mediator (if they do not want to also randomise a mediator fully).
